Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-06-17 paper/paper_R2.tex
revisions complete

  pdfauthor={Timothy J. Miller1; Kiersten Curti; Alexander C. Hansell},

@@

\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kiersten Curti \and Alexander C. Hansell}
\date{18 March, 2025}

@@

A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

WHAM is an R package developed and maintained at NOAA's Northeast Fisheries Science Center {[}\url{https://timjmiller.github.io/wham}{]}. WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects to models with several time and age varying process errors and effects of environmental covariates on various population parameters. Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al. 2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has only allowed one stock (component) and a single region and without seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al. 2020). We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0. Many of these new configuration options can also be useful when modeling a single stock and region.

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023). Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment. The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016). Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009). Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al. 2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al. 2016). Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process. However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al. 2021). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points. We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks each with their own movement, harvest, and natural mortality. Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.

@@

Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987; Commenges 1999; Andersen and Keiding 2002). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\) on January 1. Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).

WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would be multinomial distributed as in a model for tagging data where fates of individual tagged fish are assumed independent. The PTMs determine the expected numbers 1) in each state on January 1 of year \(t+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(t\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.

@@

When there is only one region,
\begin{equation}\label{eq:ptm_1_region}
\mathbf{P}_{y,a,i} = 
  \begin{bmatrix}
     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
     0 & \mathbf{I}_{H} & 0\\
     0 & 0 & 1
  \end{bmatrix}
\end{equation}
where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{y,a,i,f}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{y,a,i,f}\) is the total mortality rate.

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons. So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{y,a,i,f}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(\delta_i\) and all other elements will be zero. Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\)

When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates. We obtain the PTM over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
\begin{equation*}
\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
\end{equation*}
The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region. Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero. For two regions and one fleet operating in each region:
\begin{equation*}
 \mathbf{A}_{y,a,i} = \begin{bmatrix}
 a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
 \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0
 \end{bmatrix}
\end{equation*}
where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\). When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq. \ref{eq:ptm_1_region}.

@@

Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior. The annual time step can be divided into \(K\) seasons and the interval size \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval. Under the Markov assumption, the PTM of surviving and moving and dying over \(K\) intervals \(\delta_1,\ldots, \delta_K\) (i.e., the entire year) is just the product of the PTMs for each interval:
\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]

For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(\delta_j\), the fraction of time in season \(j\) is
\[\delta_{s,j} = t_s-\sum^{j-1}_{i=0}\delta_i
\]
and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
\begin{equation}\label{eq:ptm_spawn}
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{s,j}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{s,j}).
\end{equation}
Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(\delta_j\) the proportions in each state at the time of the observation is
\begin{equation} \label{eq:ptm_index} 
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{j-1}, \delta_{m,j}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{j-1}_{i=1}\mathbf{P}_{y,a,i}(\delta_i)\right]\mathbf{P}_{y,a,j}(\delta_{m,j}).
\end{equation}

@@

When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\). So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020). If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function. However, models can configure spawning individuals to occur in other regions at the time of spawning. Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
\begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*}
where\\
\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq. \ref{eq:ptm_spawn}).

>>>>

  pdfauthor={Timothy J. Miller1; Kiersten L. Curti; Alexander C. Hansell},

@@

\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kiersten L. Curti \and Alexander C. Hansell}
\date{16 June, 2025}

@@

State-space age-structured stock assessment models can be used to estimate time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024a).

WHAM is an R package developed and maintained by scientists at NOAA's Northeast Fisheries Science Center (\url{https://timjmiller.github.io/wham}). WHAM can be configured to fit a wide range of age-structured models from traditional statistical catch-at-age models without any random effects, to models with several time and age varying process errors and effects of environmental covariates on various population parameters. Like SAM, WHAM models are built using the Template Model Builder package (TMB, Kristensen et al. 2016) which provides a computationally efficient means of fitting an extremely wide class of models with random effects. WHAM has undergone active development since its creation and includes random effects options for recruitment, inter-annual transitions of numbers at age (hereafter referred to as ``apparent survival''), fishery and index selectivity, natural mortality, and catchability.

However, WHAM has only allowed consideration of a single stock and region, and no seasonal changes in stock dynamics. Using such models for stocks that have subcomponents with varying seasonal movement can provide incorrect inferences and poor management advice (Ying et al. 2011; Cao et al. 2014; Bosley et al. 2022). Furthermore, the ability to account for spatial structure and model multiple stocks are also important features of leading-edge assessment modeling frameworks (Punt et al. 2020). We describe here the implementation of these features and other extensions since Stock and Miller (2021) in WHAM version 2.0. Many of these new configuration options can also be useful when modeling a single stock and region.

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (BSB, \textit{Centropristis striata}; NEFSC 2023), which is a high profile stock in the Northeast U.S. and important to both commercial fishing fleets and recreational anglers. BSB are targeted from Maine to North Carolina in both state and federal waters. Past stock assessments have estimated that the stock is healthy (not overfished and overfishing not occurring). Those assessments assumed two stock components (north and south), each assessed with a separate statistical catch at age model, with results subsequently combined to inform management. However, the assessments exhibited strong retrospective patterns. The stock was split into two regional components to account for spatial dynamics and improve model diagnostics (ASMFC 2016). Additionally, BSB have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009). BSB distribution has been linked to warming waters on the northwest shelf (Bell et al. 2015) and it is hypothesized that the distribution and productivity of the BSB stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al. 2016). Thus, a focus of the BSB research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement, and explore environmental drivers of population dynamics (NEFSC 2023). We denote the fish in the north as separate stock components rather than stocks because both components are managed as a single stock, but recognizing differences in migration between fish originating in the two areas.

To illustrate the usage of WHAM 2.0 we apply it to the two stock components of BSB off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process (NEFSC 2023). However, here we evaluate evidence for alternative hypotheses of temporal variation and effects of bottom temperature, specifically, on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components and regions, further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al. 2021). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires assumptions on how they are treated in projections and reference points. We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Many of the options and equations of WHAM version 2.0 are the same as those in Stock and Miller (2021), so we will only describe extensions and differences that have occurred since their first description of WHAM. The new version of WHAM can model multiple stocks, each with their own movement, harvest, and natural mortality. Much of the description below is for a specific stock \(s\), but, for simplicity, this subscript is implicit except when necessary.

@@

Because individuals may be alive in one of several regions or harvested by one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987; Commenges 1999; Andersen and Keiding 2002). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval with duration \(\delta\) (columns), conditional on being in each of those states at the beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\). Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to one (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix},
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R \times 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the \(n_F\times n_F\) identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is one (no zombies allowed).

WHAM uses these PTMs to model abundance proportions in each state rather than true probabilities where numbers in each state would have a multinomial distribution as in a model for tagging data where fates of individual tagged fish are assumed independent. The PTMs determine the expected numbers 1) in each state on January 1 of year \(y+1\) at age \(a+1\) given the abundances at age \(a\) on January 1 of year \(y\), 2) captured over the year in each fleet, 3) available to each index, and 4) alive at the time and in the region where spawning occurs.

@@

When there is only one region,
\begin{equation}\label{eq:ptm_1_region}
\mathbf{P}_{y,a,i} = 
  \begin{bmatrix}
     S_{y,a,i} & \mathbf{H}_{y,a,i} & D_{y,a,i} \\
     0 & \mathbf{I}_{H} & 0\\
     0 & 0 & 1
  \end{bmatrix},
\end{equation}
where \(S_{y,a,i} = e^{-Z_{y,a,i}\delta_i}\), \(\mathbf{H}_{y,a,i}\) is a 1 x \(n_F\) matrix with elements for each fleet \(f\): \(\frac{F_{f,y,a,i}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), \(D_{y,a,i} = \frac{M_{y,a}}{Z_{y,a,i}}\left(1 - e^{-Z_{y,a,i}\delta_i}\right)\), and \(Z_{y,a,i} = M_{y,a} + \sum^{n_F}_{f=1} F_{f,y,a,i}\) is the total mortality rate.

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i},
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of proportions moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

WHAM assumes each fleet \(f\) can harvest in only one region (\(r_f\)) during specified seasons. So, for each fleet \(f\), row \(r_f\) and column \(f\) of \(\mathbf{H}_{y,a,i}\) will be \(F_{f,y,a,i}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\) when fleet \(f\) is harvesting during the interval \(i\) and all other elements will be zero. Row \(r\) of the single-column matrix \(\mathbf{D}_{y,a,i}\) is \(M_{y,a,r}\left(1 - e^{-Z_{y,a,i,r}\delta_i}\right)/Z_{y,a,i,r}\). For details defining PTMs with movement and mortality occurring simultaneously see Appendix B.

@@

Seasonality can be configured to accommodate characteristics of spawning, movement, and fleet-specific behavior. The annual time step can be divided into \(K\) seasons and the interval duration \(\delta_i\) for each season \(i\) does not need to be equal to any other seasonal interval. Under the Markov assumption the PTMs for each consecutive interval are independent and the PTM of surviving, moving, and dying over \(K\) intervals \(1,\ldots, K\) (i.e., the entire year) is just the product of the PTMs for each interval:
\[ \mathbf{P}_{y,a}(\delta_1,\ldots,\delta_K) = \prod^K_{i=1}\mathbf{P}_{y,a,i}(\delta_i).\]

For a stock spawning at some fraction of the year \(0<t_s<1\) in interval \(i \in \{1,\ldots,K\}\), the fraction of time in the interval is
\[\delta_{s,i} = t_s-\sum^{i-1}_{j=0}\delta_j\]
(where \(\delta_0 = 0\)), and the PTM defining the proportions in each state at time \(t_s\) for age \(a\) is
\begin{equation}\label{eq:ptm_spawn}
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{s,i}\right) = \mathbf{P}_{y,a}\left(t_s\right) =  \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,j}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{s,i}),
\end{equation}
where \(\mathbf{P}_{y,a,0}\) is an identity matrix. Similarly, for an index \(m\) occurring at fraction of the year \(t_m\) in interval \(i\) the proportions in each state at the time of the observation is
\begin{equation} \label{eq:ptm_index} 
\mathbf{P}_{y,a}\left(\delta_1,\ldots,\delta_{i-1}, \delta_{m,i}\right) = \mathbf{P}_{y,a}\left(t_m\right) =   \left[\prod^{i-1}_{j=0}\mathbf{P}_{y,a,i}(\delta_j)\right]\mathbf{P}_{y,a,i}(\delta_{m,i}).
\end{equation}

@@

When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance for a given stock (component) in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a}^T = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)\), where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet or dead from natural mortality because no age \(a\) fish have died yet on January 1.

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\). So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM. If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function. However, models can configure individuals to occur in other regions at the time of spawning. Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
\begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1} =  f\left(\text{SSB}_{y-1}\right) + \varepsilon_{y,1,r_s},
\end{equation*}
where\\
\[\text{SSB}_{y} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,\cdot,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
and \(w_{y,a}\) and \(\text{mat}_{y,a}\) are the stock-specific mean weight at age of spawning individuals and maturity at age, respectively, and \(\mathbf{O}_{y,a,\cdot,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (column \(r_s\) of the upper-left submatrix of Eq. \ref{eq:ptm_spawn}).

As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age \(a\) in year \(y\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021). For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
\[\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.\]
The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the vector of abundance at age in year \(y\) at age \(a\) given the vector in year \(y-1\) at age \(a-1\) is
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A},
\end{equation*}
where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
